---
title: "Isolate Functional Report"
subtitle: "`r params$sample_name`"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_width: 10
    fig_height: 6
    latex_engine: xelatex
    keep_tex: false
geometry: margin=0.75in
fontsize: 10pt
header-includes:
  - \usepackage{booktabs}
  - \usepackage{array}
  - \usepackage{xcolor}
  - \usepackage{caption}
  - \captionsetup[table]{skip=4pt}
params:
  report: NULL
  sample_name: "Isolate"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 10,
  fig.height = 6
)

library(stringr)
library(ggplot2)
library(patchwork)
library(knitr)
library(scales)

rpt <- params$report
fd  <- rpt$featuredata

# ── Helpers ──────────────────────────────────────────────────────────────────

# Safe base-R column subset — immune to MASS/dplyr conflicts
safe_select <- function(df, cols) {
  cols <- intersect(cols, colnames(df))
  df[, cols, drop = FALSE]
}

# Plain booktabs kable — no kableExtra, no tabu
pkable <- function(df, col.names = NA, caption = NULL, row.names = FALSE, ...) {
  if (identical(col.names, NA)) col.names <- colnames(df)
  knitr::kable(df,
               format      = "latex",
               booktabs    = TRUE,
               col.names   = col.names,
               caption     = caption,
               row.names   = row.names,
               linesep     = "",
               ...)
}

# Chunk large data frames and print one table per chunk
chunked_pkable <- function(df, n = 45, col.names = NA, ...) {
  if (is.null(df) || nrow(df) == 0) { cat("> No data.\n"); return(invisible(NULL)) }
  if (identical(col.names, NA)) col.names <- colnames(df)
  chunks <- split(df, ceiling(seq_len(nrow(df)) / n))
  for (ch in chunks) {
    print(pkable(ch, col.names = col.names, ...))
    cat("\n\n")
  }
}
```

\newpage

# Overview

```{r overview}
n_genes    <- nrow(fd)
n_cds      <- sum(fd$FeatType == "CDS", na.rm = TRUE)
n_with_ec  <- sum(!is.na(fd$ECNumber) & fd$ECNumber != "none" & fd$ECNumber != "")
n_with_go  <- sum(!is.na(fd$GO)       & fd$GO       != "none" & fd$GO       != "")
n_with_ipr <- sum(!is.na(fd$Interpro) & fd$Interpro != "none" & fd$Interpro != "")
n_hypo     <- sum(str_detect(tolower(fd$Product), "hypothetical"), na.rm = TRUE)
total_bp   <- sum(fd$NumBases, na.rm = TRUE)

overview_df <- data.frame(
  Metric = c("Total features", "CDS features", "Total bases annotated (bp)",
             "Genes with EC number", "Genes with GO term",
             "Genes with InterPro hit", "Hypothetical proteins"),
  Value  = c(n_genes, n_cds, format(total_bp, big.mark = ","),
             n_with_ec, n_with_go, n_with_ipr, n_hypo),
  stringsAsFactors = FALSE
)

pkable(overview_df, col.names = c("Metric", "Value"),
       caption = "Sample annotation summary")
```

\newpage

# Functional Landscape

## Gene categories (keyword mining)

```{r category_plot, fig.height=5}
product_hits <- rpt$product_hits

cat_counts <- do.call(rbind, lapply(names(product_hits), function(cat) {
  data.frame(Category = cat, n = nrow(product_hits[[cat]]), stringsAsFactors = FALSE)
}))
cat_counts <- cat_counts[cat_counts$n > 0, ]
cat_counts <- cat_counts[order(-cat_counts$n), ]
cat_counts$Category <- str_to_title(str_replace_all(cat_counts$Category, "_", " "))
cat_counts$Category <- factor(cat_counts$Category, levels = cat_counts$Category)

ggplot(cat_counts, aes(x = Category, y = n, fill = n)) +
  geom_col(show.legend = FALSE) +
  scale_fill_viridis_c(option = "D") +
  coord_flip() +
  labs(title = "Functional gene categories",
       subtitle = "Keyword mining of PROKKA Product descriptions",
       x = NULL, y = "Gene count") +
  theme_minimal(base_size = 11)
```

```{r category_count_table}
pkable(cat_counts, col.names = c("Functional Category", "Gene Count"),
       caption = "Gene counts per functional category")
```

\newpage

## Gene lists per category

```{r category_gene_lists, results='asis'}
for (cat in names(product_hits)) {
  df <- product_hits[[cat]]
  if (is.null(df) || !is.data.frame(df) || nrow(df) == 0) next
  cat(sprintf("\n### %s\n\n",
              str_to_title(str_replace_all(cat, "_", " "))))
  out <- safe_select(df, c("Feature", "Product", "ECNumber"))
  out <- unique(out)
  if ("Product" %in% colnames(out))
    out$Product <- str_trunc(out$Product, 50)
  chunked_pkable(out, n = 45, row.names = FALSE)
  cat("\n")
}
```

\newpage

# KEGG Pathway Analysis

```{r kegg_check}
has_kegg <- !is.null(rpt$kegg) &&
            !is.null(rpt$kegg$completeness) &&
            nrow(rpt$kegg$completeness) > 0
```

```{r kegg_plot, eval=has_kegg, fig.height=6}
comp   <- rpt$kegg$completeness
top_pw <- head(comp[order(-comp$completeness), ], 25)
top_pw$Name <- str_trunc(top_pw$Name, 45)
top_pw$Name <- factor(top_pw$Name, levels = rev(top_pw$Name))
top_pw$fill_col <- cut(top_pw$completeness,
                        breaks = c(0, 0.5, 0.75, 1.001),
                        labels = c("Low (<50%)", "Medium (50-74%)", "High (>=75%)"),
                        right  = FALSE)

ggplot(top_pw, aes(x = Name, y = completeness, fill = fill_col)) +
  geom_col() +
  scale_fill_manual(
    values = c("High (>=75%)"    = "#2ecc71",
               "Medium (50-74%)" = "#f39c12",
               "Low (<50%)"      = "#e74c3c"),
    name = "Completeness") +
  geom_hline(yintercept = 0.50, linetype = "dashed", linewidth = 0.5) +
  geom_hline(yintercept = 0.75, linetype = "dashed",
             color = "darkgreen", linewidth = 0.5) +
  coord_flip() +
  scale_y_continuous(labels = percent) +
  labs(title = "KEGG Pathway Completeness (top 25)",
       x = NULL, y = "Completeness") +
  theme_minimal(base_size = 11)
```

## Pathways $\geq$ 50% complete

```{r kegg_table_above, eval=has_kegg}
above <- rpt$kegg$above_threshold
above$Name         <- str_trunc(above$Name, 46)
above$completeness <- percent(above$completeness)
out <- safe_select(above, c("Pathway", "Name", "n_found", "n_total", "completeness"))
chunked_pkable(out, n = 45,
               col.names = c("KEGG ID", "Pathway Name",
                              "ECs found", "ECs total", "Completeness"),
               caption = "KEGG pathways with >= 50% EC completeness")
```

## Full pathway completeness table

```{r kegg_table_full, eval=has_kegg}
full <- rpt$kegg$completeness
full$Name         <- str_trunc(full$Name, 46)
full$completeness <- percent(full$completeness)
out <- safe_select(full, c("Pathway", "Name", "n_found", "n_total", "completeness"))
chunked_pkable(out, n = 45,
               col.names = c("KEGG ID", "Pathway Name",
                              "ECs found", "ECs total", "Completeness"),
               caption = "All KEGG pathways (sorted by completeness)")
```

\newpage

# GO Term Analysis

```{r go_check}
has_go <- !is.null(rpt$go) &&
          !is.null(rpt$go$go_info) &&
          nrow(rpt$go$go_info) > 0
```

```{r go_plot, eval=has_go, fig.height=4}
go_info <- rpt$go$go_info
go_info <- go_info[!is.na(go_info$namespace), ]
go_info$namespace <- str_to_title(str_replace_all(go_info$namespace, "_", " "))
ns_counts <- as.data.frame(table(go_info$namespace), stringsAsFactors = FALSE)
colnames(ns_counts) <- c("Namespace", "Count")

ggplot(ns_counts, aes(x = reorder(Namespace, Count), y = Count, fill = Namespace)) +
  geom_col(show.legend = FALSE) +
  scale_fill_brewer(palette = "Set2") +
  coord_flip() +
  labs(title = "GO Term Distribution by Namespace", x = NULL, y = "Count") +
  theme_minimal(base_size = 11)
```

```{r go_tables, eval=has_go, results='asis'}
go_info_raw <- rpt$go$go_info
go_info_raw <- go_info_raw[!is.na(go_info_raw$name), ]

ns_map <- c(
  "biological_process" = "Biological Process",
  "molecular_function" = "Molecular Function",
  "cellular_component" = "Cellular Component"
)

for (ns in names(ns_map)) {
  sub <- go_info_raw[go_info_raw$namespace == ns, ]
  if (is.null(sub) || nrow(sub) == 0) next
  cat(sprintf("\n## %s\n\n", ns_map[ns]))
  sub <- sub[order(sub$GO_id), ]
  out <- safe_select(sub, c("GO_id", "name"))
  chunked_pkable(out, n = 45,
                 col.names = c("GO Accession", "Term Name"),
                 row.names = FALSE)
  cat("\n")
}
```

\newpage

# InterPro Domain Analysis

```{r ipr_check}
has_ipr <- !is.null(rpt$interpro) && nrow(rpt$interpro) > 0
```

```{r ipr_plot, eval=has_ipr, fig.height=4}
ipr_info <- rpt$interpro[!is.na(rpt$interpro$type), ]
type_counts <- as.data.frame(table(ipr_info$type), stringsAsFactors = FALSE)
colnames(type_counts) <- c("Type", "Count")

ggplot(type_counts, aes(x = reorder(Type, Count), y = Count, fill = Type)) +
  geom_col(show.legend = FALSE) +
  scale_fill_brewer(palette = "Set1") +
  coord_flip() +
  labs(title = "InterPro Hit Types", x = NULL, y = "Count") +
  theme_minimal(base_size = 11)
```

```{r ipr_table, eval=has_ipr}
ipr <- rpt$interpro[!is.na(rpt$interpro$name), ]
ipr <- ipr[order(ipr$type, ipr$IPR_id), ]
chunked_pkable(ipr, n = 45,
               col.names = c("IPR Accession", "Domain/Family Name", "Type"),
               row.names = FALSE)
```

\newpage

# Targeted Pathway Probes

```{r probes, results='asis'}
probes <- rpt$probes

for (i in seq_along(probes)) {
  pr <- probes[[i]]
  cat(sprintf("\n## %s\n\n", pr$label))
  cat(pr$summary, "\n\n")
  if (!is.null(pr$genes) && nrow(pr$genes) > 0) {
    out <- safe_select(pr$genes, c("Feature", "Product", "ECNumber", "NumBases"))
    out <- unique(out)
    if ("Product" %in% colnames(out))
      out$Product <- str_trunc(out$Product, 50)
    chunked_pkable(out, n = 40, row.names = FALSE)
  } else {
    cat("> No genes matched this probe.\n")
  }
  cat("\n")
}
```

\newpage

# Hypothetical Proteins with InterPro Hits

Hypothetical proteins carrying an InterPro annotation reveal functional
information despite their uninformative Product names.

```{r hypo_ipr, eval=has_ipr}
hypo <- fd[
  str_detect(tolower(fd$Product), "hypothetical") &
  !is.na(fd$Interpro) & fd$Interpro != "none" & fd$Interpro != "",
]

if (nrow(hypo) > 0) {
  hypo$First_IPR <- str_extract(hypo$Interpro, "IPR\\d+")
  hypo <- merge(hypo, rpt$interpro,
                by.x = "First_IPR", by.y = "IPR_id", all.x = TRUE)
  out <- safe_select(hypo, c("Feature", "First_IPR", "name", "type", "ECNumber"))
  out <- unique(out)
  chunked_pkable(out, n = 40,
                 col.names = c("Feature", "IPR Accession",
                                "Domain Name", "Type", "EC Number"),
                 row.names = FALSE)
} else {
  cat("> No hypothetical proteins with InterPro hits found.\n")
}
```

\newpage

# Session Info

```{r session_info, echo=TRUE}
sessionInfo()
```